An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2022-12-27 article/article.tex
finished chapter 2

\newcommand{\simulator}{g}
\newcommand{\region}{B_{d,\epsilon}}
\newcommand{\indicator}[1]{\mathbb{1}_{#1}}
\newcommand{\regioni}{B_{d,\epsilon}^i}
\newcommand{\data}{\mathbf{y_0}}
\newcommand{\accregioni}{C^i_{\epsilon}}
\newcommand{\accregionihat}{\hat{C}^i_{\epsilon}}

@@

An implicit or simulator-based model is a parameterized stochastic
data generating mechanism. The key characteristic of simulator-based
models is that we can sample data points, but we cannot evaluate the
likelihood. Formally, a simulator-based model is a parameterized
family of probability density functions
\(\{ p(\yb|\thetab)\}_{\thetab}\) whose closed-form is either unknown
or computationally intractable. In these scenarios, we can only a
access the simulator \( m_r(\thetab) \), i.e., the black-box mechanism
(computer code) that generates samples \(\yb\) in a stochastic manner
from a set of parameters \(\thetab\). We denote the process of
obtaining samples from the simulator with
\( m_r(\thetab) \rightarrow \yb \). As shown by~\citet{Meeds2015}, it
is feasible to isolate the randomness of the simulator by introducing
a set of nuisance random variables denoted by \(\ub \sim
p(\ub)\). Therefore, for a specific tuple \((\thetab, \ub)\) the
simulator becomes a deterministic mapping \(g\), such that
\(\yb=\simulator(\thetab,\ub)\). In terms of computer code, the
stochasticity of a random process is governed by the global
seed. Although each software may handle the randomness in slightly
different ways\footnote{For example, at
  \pkg{Numpy}~\citet{harris2020array}, the pseudo-random number
  generation is based on a global state, whereas, in
  \pkg{JAX}~\citet{jax2018github} random functions consume a key that
  is passed as parameter. }, in all cases, freezing the initial seed
to a specific integer converts the simulation to a deterministic piece
of code.

@@

Equation~\ref{eq:likelihood} highlights the main source of
intractability; computing
\(\Pr(\simulator(\thetab,\ub) \in \region(\data)) | \thetab\) as the
fraction of samples that lie inside the \(\epsilon\) region around
\(\data\) is computationally infeasible in the limit of
\(\epsilon \to 0\). Hence, the constrainted is relaxed to
\(\epsilon > 0\), which leads to the approximate likelihood:

@@

As argued by \citet{Ikonomov2019}, these derivations provide a unique
perspective for likelihood-free inference by shifting the focus onto
the geometry of the acceptance regions \(\accregioni\). Indeed, the
task of approximating the likelihood and the posterior becomes a task
of characterising the sets \(\accregioni\). OMC by \citet{Meeds2015}
assumes that the distance \(d\) is the Euclidean distance
\(||\cdot||_2\) between summary statistics \(\Phi\) of the observed
and generated data, and that the \(\accregioni\) can be well
approximated by infinitesimally small ellipses. These assumptions lead
to an approximation of the posterior in terms of weighted samples
\(\thetab_i^*\) that achieve the smallest distance between observed
and simulated data for each realisation \(\ub_i \sim p(\ub)\), i.e.,\
\begin{equation} \label{eq:omc_opt_prob}
\thetab_i^* = \argmin_{\thetab} ||\Phi(\data)-\Phi(\simulator(\thetab, \ub_i))||_2  , \quad \ub_i \sim p(\ub).
\end{equation}
The weighting for each \(\thetab_i^*\) is proportional to the prior
density at \(\thetab_i^*\) and inversely proportional to the determinant
of the Jacobian matrix of the summary statistics at \(\thetab_i^*\). For
further details on OMC we refer the reader to \citep{Meeds2015,
  Ikonomov2019}.

@@

\citet{Ikonomov2019} showed that considering infinitesimally small
ellipses can lead to highly overconfident posteriors. We refer the
reader to their paper for the technical details and conditions for
this issue to occur. Intuitively, it happens because the weights in
OMC are only computed from information at \(\thetab_i^*\), and using
only local information can be misleading. For example if the curvature
of \(||\Phi(\data)-\Phi(\simulator(\thetab, \ub_i))||_2\) at
\(\thetab_i^*\) is nearly flat, the curvature alone may wrongly
indicate that \(\accregioni\) is much larger than it actually is. In
our software package we implement the robust generalisation of OMC by
\citet{Ikonomov2019} that resolves this issue.

ROMC, firstly, approximates the acceptance regions \(\accregioni\),
and defines proposal distributions \(q_i(\thetab)\) on them. The
proposal distributions are used for generating posterior samples
\(\thetab_{ij} \sim q_i\). The samples are assigned (importance)
weights \(w_{ij}\) that compensate for using the proposal
distributions \(q_i(\thetab)\) and not the prior \(p(\thetab)\),
\begin{equation}
  w_{ij} = \frac{\indicator{\accregioni}(\thetab_{ij}) p(\thetab_{ij})}{q(\thetab_{ij})}.
  \label{eq:sampling}
\end{equation}
Given the weighted samples, any expectation
\(\E_{p(\thetab|\data)}[h(\thetab)]\) of some function \(h(\thetab)\), can be approximated as
\begin{equation} \label{eq:expectation}
  \E_{p(\thetab|\data)}[h(\thetab)] \approx \frac{\sum_{ij} w_{ij} h(\thetab_{ij})}{\sum_{ij} w_{ij}}
\end{equation}
\citet{Ikonomov2019} considered uniform distributions as proposal
distributions so that the main task is to approximate the acceptance
regions \(\accregioni\) and to represent them so that uniform sampling
is easy. The approximation of the acceptance regions contains two
compulsory and one optional step: (1) solving the optimisation
problems as in OMC, (2) constructing bounding boxes around
\(\accregioni\) and optionally, (3) refining the approximation via a
surrogate model of the distance.

\subsubsection*{Solving the deterministic optimisation problems}
For each set of nuisance variables \(\ub_i, i = \{1,2,\ldots,n_1 \}\),
we search for a point \(\thetab^*_i\) such that
\(d(\simulator(\thetab^*_i,\ub_i), \data) \le \epsilon\). In general,
\(d\) can be any valid distance, but for the rest of the paper we
consider \(d\) as the squared Euclidean distance. For notational
convenience, we denote \(d(\simulator(\thetab,\ub_i), \data)\) as
\(d_i(\thetab)\).  Obtaining \(\theta_i^*\) involves solving the
following optimisation problem:
\begin{subequations}
\begin{alignat}{2}
  &\!\min_{\thetab}        && d_i(\thetab) \label{eq:optProb}\\
  &\text{subject to} & \quad& d_i(\thetab) \leq \epsilon
\end{alignat}
\end{subequations}
%
The optimisation problem can be treated as unconstrained, accepting
the optimal point
\(\thetab_i^* = \text{argmin}_{\thetab} d_i(\thetab)\) only if
\(d_i(\thetab_i^*) \le \epsilon\). If \(d_i(\thetab)\) is
differentiable any gradient-based optimizer can be used
for~\ref{eq:optProb}. The gradients \(\nabla_{\thetab} d_i(\thetab)\)
can be either provided in closed form or approximated by finite
differences. In case \(d_i\) is not differentiable, Bayesian
Optimisation~\citep{Shahriari2016} provides an alternative
approach. In this scenario, apart from obtaining an optimal
\(\thetab_i^* \), a surrogate model \(\hat{d}_i(\thetab)\) of the
distance function \(d_i(\thetab)\) is also automatically obtained;
\(\hat{d}_i\) can then substitute the actual distance function in
downstream steps of the algorithms, with possible computational gains
especially if evaluating the actual distance \(d_i(\thetab)\) is
expensive.

\subsubsection*{Estimating the acceptance regions}
The acceptance region \(\accregioni\) is approximated by a bounding
box \(\accregionihat\). Ideally, we want the bounding box to be as
tight as possible to \(\accregioni\) to ensure high acceptance rate in
the importance sampling, but big enough for not discarding valid
parts. The bounding boxes are built in two steps. First, we define
their axes \(\mathbf{v}_m\), \(m = \{1, \ldots, D\}\) based on the
(estimated) curvature of the distance at \(\thetab_i^*\). Second, we
determine the size of the box via a one-dimensional line-search method
along each axis, see Algorithm~\ref{alg:region_construction} for the
details. After the bounding boxes construction, a uniform distribution
\(q_i\) is defined on each bounding box, and is used as the proposal
region for importance sampling.

\subsubsection*{Refining the estimate via a local surrogate model (optional)}
When computing the weight \(w_{ij}\) in \eqref{eq:sampling}, we need
to check whether the samples \(\thetab_{ij} \sim q_i\) lie inside the
acceptance region \(\accregioni\). This can be considered to be a
safety-mechanism that corrects for any inaccuracies in the
construction of \(\accregionihat\) above. However, this check involves
evaluting the distance function \(d_i(\thetab_{ij})\), which can be
expensive if the model is complex. \citet{Ikonomov2019} thus proposed
to fit a surrogate model \(\tilde{d}_i(\thetab)\) of the distance
function \(d_i(\thetab)\), on data points that lie inside
\(\accregionihat\). They used a simple quadratic model whilst other
regression models are, in principle, possible too. The advantage of
using a quadratic model is that it has ellipsoidal isocontours, which
thus naturally allowed \citet{Ikonomov2019} to replace the bounding
box approximation of \(\accregioni\) with a tigher-fitting ellipsoidal
approximation.\footnote{The difference to the infinitesimal
  ellipsoidal model in OMC is the estimation procedure: OMC uses
  information at \(\thetab_i^*\) whilst, here, information in
  \(\accregionihat\) is used, which results in a more stable fit.}

The training data for the quadratic model is obtained by sampling
\(\thetab_{ij} \sim q_i\) and accessing the distances
\(d_i(\thetab_{ij})\). The generation of the training data adds an
extra computational cost, but leads to a significant speed-up when
evaluating the weights \(w_{ij}\). Moreover, the extra cost is largely
eliminated if Bayesian Optimisation with a Gaussian process (GP)
surrogate model \(\hat{d}_i(\thetab)\) was used to obtain
\(\thetab_i^*\) in the first step. In this case, we can use
\(\hat{d}_i(\thetab)\) instead of \(d_i(\thetab)\) to generate the
training data. This essentially replaces the global GP model with a
simpler local quadratic model which is typically more robust.

\subsection[Engine for likelihood-free inference (ELFI)]{\pkg{Engine for likelihood-free inference (ELFI)}}
\label{subsec:ELFI}
\input{latex_files/ch2/2-4-ELFI.tex}

>>>>

\newcommand{\simulator}{g}
\newcommand{\region}{B_{d,\epsilon}}
\newcommand{\indicator}[1]{\mathbb{1}_{#1}}
\newcommand{\regioni}{B_{d,\epsilon}^i}
\newcommand{\data}{\mathbf{y_0}}
\newcommand{\accregioni}{C^i_{\epsilon}}
\newcommand{\accregionitheta}{C^i_{\epsilon, \theta_i^*}}
\newcommand{\accregionihat}{\hat{C}^i_{\epsilon}}

@@

An implicit or simulator-based model is a parameterized stochastic
data generating mechanism. The key characteristic of simulator-based
models is that we can sample data points, but we cannot evaluate the
likelihood. Formally, a simulator-based model is a parameterized
family of probability density functions
\(\{ p(\yb|\thetab)\}_{\thetab}\) whose closed-form is either unknown
or computationally intractable. In these scenarios, we can only a
access the simulator \( m_r(\thetab) \), i.e., the black-box mechanism
(computer code) that generates samples \(\yb\) in a stochastic manner
from a set of parameters \(\thetab\). We denote the process of
obtaining samples from the simulator with
\( m_r(\thetab) \rightarrow \yb \). As shown by~\citet{Meeds2015}, it
is feasible to isolate the randomness of the simulator by introducing
a set of nuisance random variables denoted by \(\ub \sim
p(\ub)\). Therefore, for a specific tuple \((\thetab, \ub)\) the
simulator becomes a deterministic mapping \(g\), such that
\(\yb=\simulator(\thetab,\ub)\). In terms of computer code, the
randomness of a random process is governed by the global
seed. Although each software may handle the randomness in slightly
different ways\footnote{For example, at
  \pkg{Numpy}~\citet{harris2020array}, the pseudo-random number
  generation is based on a global state, whereas, in
  \pkg{JAX}~\citet{jax2018github} random functions consume a key that
  is passed as parameter. }, in all cases, freezing the initial seed
to a specific integer converts the simulation to a deterministic piece
of code.

@@

Equation~\ref{eq:likelihood} highlights the main source of
intractability; computing
\(\Pr(\simulator(\thetab,\ub) \in \region(\data)) | \thetab\) as the
fraction of samples that lie inside the \(\epsilon\) region around
\(\data\) is computationally infeasible in the limit of
\(\epsilon \to 0\). Hence, the constraint is relaxed to
\(\epsilon > 0\), which leads to the approximate likelihood:

@@

As argued by \citet{Ikonomov2019}, these derivations provide a unique
perspective for likelihood-free inference by shifting the focus onto
the geometry of the acceptance regions \(\accregioni\). Indeed, the
task of approximating the likelihood and the posterior becomes a task
of characterizing the sets \(\accregioni\). OMC by \citet{Meeds2015}
assumes that the distance \(d\) is the Euclidean distance
\(||\cdot||_2\) between summary statistics \(\Phi\) of the observed
and generated data, and that the \(\accregioni\) can be well
approximated by infinitesimally small ellipses. These assumptions lead
to an approximation of the posterior in terms of weighted samples
\(\thetab_i^*\) that achieve the smallest distance between observed
and simulated data for each realization \(\ub_i \sim p(\ub)\), i.e.,\
\begin{equation} \label{eq:omc_opt_prob}
\thetab_i^* = \argmin_{\thetab} ||\Phi(\data)-\Phi(\simulator(\thetab, \ub_i))||_2  , \quad \ub_i \sim p(\ub).
\end{equation}
The weighting for each \(\thetab_i^*\) is proportional to the prior
density at \(\thetab_i^*\) and inversely proportional to the determinant
of the Jacobian matrix of the summary statistics at \(\thetab_i^*\). For
further details on OMC we refer the reader to \citep{Meeds2015,
  Ikonomov2019}.

@@

\citet{Ikonomov2019} showed that considering infinitesimally small
ellipses can lead to highly overconfident posteriors. We refer the
reader to their paper for the technical details and conditions for
this issue to occur. Intuitively, it happens because the weights in
OMC are only computed from information at \(\thetab_i^*\), and using
only local information can be misleading. For example, if the
curvature of \(||\Phi(\data)-\Phi(\simulator(\thetab, \ub_i))||_2\) at
\(\thetab_i^*\) is nearly flat, it may wrongly indicate that
\(\accregioni\) is much larger than it actually is. In our software
package we implement the robust generalization of OMC by
\citet{Ikonomov2019} that resolves this issue.

ROMC approximates the acceptance regions \(\accregioni\) with
\(D\)-dimensional bounding boxes \(\accregionihat\). A uniform
distribution, \(q_i(\thetab)\), is defined on top of each bounding
box. Each \(q_i(\thetab)\) serves a proposal distribution for
generating posterior samples \(\thetab_{ij} \sim q_i\). The samples
get an (importance) weight \(w_{ij}\) that compensate for using the
proposal distributions \(q_i(\thetab)\) instead of the prior
\(p(\thetab)\):

\begin{equation}
  \label{eq:sampling}
  w_{ij} = \indicator{\accregioni}(\thetab_{ij}) \frac{ p(\thetab_{ij})}{q(\thetab_{ij})}.
\end{equation}

Given the weighted samples, any expectation
\(\E_{p(\thetab|\data)}[h(\thetab)]\) of some function \(h(\thetab)\), can be approximated as
\begin{equation} \label{eq:expectation}
  \E_{p(\thetab|\data)}[h(\thetab)] \approx \frac{\sum_{ij} w_{ij} h(\thetab_{ij})}{\sum_{ij} w_{ij}}
\end{equation}
\citet{Ikonomov2019} considered uniform distributions as proposal
distributions so that the main task is to approximate the acceptance
regions \(\accregioni\) and to represent them so that uniform sampling
is easy. The approximation of the acceptance regions contains two
compulsory and one optional step: (i) solving the optimization
problems as in OMC, (ii) constructing bounding boxes around
\(\accregioni\) and, optionally, (iii) refining the approximation via a
surrogate model of the distance.

\subsubsection*{(i) Solving the deterministic optimization problems}
For each set of nuisance variables \(\ub_i, i = \{1,2,\ldots,n_1 \}\),
we search for a point \(\thetab^*_i\) such that
\(d(\simulator(\thetab^*_i,\ub_i), \data) \le \epsilon\). In
principle, \(d(\cdot)\) can refer to any valid distance function. For
the rest of the paper we consider \(d(\cdot)\) as the squared
Euclidean distance, as in~\citet{Ikonomov2019}. For easier notation,
we use \(d_i(\thetab)\) to refer to
\(d(\simulator(\thetab,\ub_i), \data)\). We search for \(\theta_i^*\)
solving the following optimization problem:

\begin{equation}
  \label{eq:optProb}
  \thetab_i^* = \!\argmin_{\thetab} d_i(\thetab)
\end{equation}
%
The solution is accepted only if it satisfies the constraint
\(d_i(\thetab_i^*) \le \epsilon\). If not, there is no point into
building an acceptance region around \(\accregioni\), so
\(\thetab_i^*\) is discarded. There are two proposed options for
solving Equation~\ref{eq:optProb} depending on the differentiability
of \(d_i(\thetab)\). If \(d_i(\thetab)\) is differentiable any
gradient-based optimizer can be used for
Equation~\ref{eq:optProb}. The gradients
\(\nabla_{\thetab} d_i(\thetab)\) can be either provided in closed
form or approximated by finite differences. If \(d_i\) is not
differentiable, Bayesian Optimization~\citep{Shahriari2016} can be
used instead. In this scenario, apart from obtaining an optimal
\(\thetab_i^* \), we can also automatically build a surrogate model
\(\hat{d}_i(\thetab)\) of the distance function \(d_i(\thetab)\). The
surrogate model \(\hat{d}_i\) can then substitute the actual distance
function in downstream steps of the algorithms, with possible
computational gains especially in cases where evaluating the actual
distance \(d_i(\thetab)\) is expensive.

\subsubsection*{(ii) Estimating the acceptance regions}

Each acceptance region \(\accregioni\) is approximated by a bounding
box \(\accregionihat\). The acceptance regions \(\accregioni\) can
contain any number of disjoint subsets in the \(D\)-dimensional space
and any of these subsets can take any arbitrary shape. First, since
the bounding boxes are built around \(\thetab_i^*\), we focus only on
the connected subset of \(\accregioni\) that contains \(\thetab_i^*\),
which we denote as \(\accregionitheta\). Second, we want to ensure
that the bounding box \(\accregionihat\) is big enough to contain on
its interior all the volume of \(\accregionitheta\). Third, we want to
be as tight as possible to \(\accregionihat\) to ensure high
acceptance rate on the importance sampling step that will
follow. Therefore, the bounding boxes are built as follows. First, we
compute their axes \(\mathbf{v}_m\), for \(m = \{1, \ldots, D\}\)
based on the (estimated) curvature of the distance at \(\thetab_i^*\),
and, second, we apply a line-search method along each axis to determine
the size of the box, see Algorithm~\ref{alg:region_construction} for
the details. After the bounding boxes construction, a uniform
distribution \(q_i\) is defined on each bounding box, and is used as
the proposal region for importance sampling.

\subsubsection*{(iii) Refining the estimate via a local surrogate model (optional)}

For computing the weight \(w_{ij}\) in Equation~\ref{eq:sampling}, we
need to check whether the samples \(\thetab_{ij}\) drawn from the
bounding boxes are inside the acceptance region \(\accregioni\). This
can be considered to be a safety-mechanism that corrects for any
inaccuracies in the construction of \(\accregionihat\) above. However,
this check involves evaluating the distance function
\(d_i(\thetab_{ij})\), which can be expensive if the model is
complex. \citet{Ikonomov2019} proposed fitting a surrogate model
\(\tilde{d}_i(\thetab)\) of the distance function \(d_i(\thetab)\), on
data points that lie inside \(\accregionihat\). In principle, any
regression model can be used as surrogate model. \citet{Ikonomov2019}
used a simple quadratic model because it has ellipsoidal isocontours,
which facilitates replacing the bounding box approximation of
\(\accregioni\) with a tighter-fitting ellipsoidal approximation.

The training data for the quadratic model is obtained by sampling
\(\thetab_{ij} \sim q_i\) and accessing the distances
\(d_i(\thetab_{ij})\). The generation of the training data adds an
extra computational cost, but leads to a significant speed-up when
evaluating the weights \(w_{ij}\). Moreover, the extra cost is largely
eliminated if Bayesian Optimization with a Gaussian process (GP)
surrogate model \(\hat{d}_i(\thetab)\) was used to obtain
\(\thetab_i^*\) in the first step. In this case, we can use
\(\hat{d}_i(\thetab)\) instead of \(d_i(\thetab)\) to generate the
training data. This essentially replaces the global GP model with a
simpler local quadratic model which is typically more robust.

\subsection[Engine for likelihood-free inference (ELFI)]{\pkg{Engine for likelihood-free inference (ELFI)}}
\label{subsec:ELFI}

\pkg{Engine for Likelihood-Free Inference (ELFI)}\footnote{Extended
  documentation can be found
  \href{https://elfi.readthedocs.io/en/latest/}{https://elfi.readthedocs.io}}~\cite{1708.00707}
is a \proglang{Python} package for LFI. We selected to implemented
ROMC in \pkg{ELFI} since it provides convenient modules for all the
fundamental components of a probabilistic model (e.g.\ prior,
simulator, summaries etc.). Furthermore, \pkg{ELFI} already supports
some recently proposed likelihood-free inference methods. \pkg{ELFI}
handles the probabilistic model as a Directed Acyclic Graph
(DAG). This functionality is based on the package
\pkg{NetworkX}~\cite{hagberg2008exploring}, which supports
general-purpose graphs. In most cases, the structure of a
likelihood-free model follows the pattern of Figure~\ref{fig:elfi};
some edges connect the prior distributions to the simulator, the
simulator is connected to the summary statistics that, in turn, lead
to the output node. Samples can be obtained from all nodes through
sequential (ancestral) sampling. \pkg{ELFI} automatically considers as
parameters of interest, i.e., those we try to infer a posterior
distribution, the ones included in the \code{elfi.Prior} class.

\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.8\textwidth]{./latex_files/images/chapter2/elfi.png}
    \end{center}
    \caption[Baseline example for creating an \pkg{ELFI} model]{Baseline example for creating an \pkg{ELFI} model. Image taken from \cite{1708.00707}}
    \label{fig:elfi}
\end{figure}

All inference methods are implemented as separate \pkg{ELFI} classes
with the two conventions:

\begin{itemize}
\item their constructor is \code{elfi.<Class name>(<output node>, *arg)};
  the initial argument is the output node of the simulator-based model
  and the rest of the arguments are parameters of the method.
\item they obtain posterior samples using \code{elfi.<Class name>.sample(*args)}
\end{itemize}
